---
title: "Birth Rates in the US from 1995 to 2017 : An Exploratory Analysis"
description: |
  Responses Submitted for Precision-Analytics Entry Exam
author:
  - name: Corey Pembleton
    url: https://data-break.netlify.com
    affiliation: Precision-Analytics
    affiliation_url: https://www.precision-analytics.ca/
date: "`r Sys.Date()`"
output: 
  distill::distill_article:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1. Load the Data
```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(naniar)
library(kableExtra)
library(DataExplorer)
```


```{r, warning=FALSE, message=FALSE}
paths <- paste0("fertility-challenge/", list.files("fertility-challenge/", pattern = "*.txt"))
Births <- map_dfr(paths, ~ read_tsv(.x))
```

Looks like on the data load some parsing errors occured in one of the columns where only one of the 10 columns is used (Notes column) - will review this later.

# 2. Exploratory Data Analysis

Can get to know the data a few ways, first, a summary of what it looks like w/ base R:

```{r}
summary(Births)
```

And to visualize this by type and oberservation, the ```DataExplorer``` package can help:

```{r, echo=FALSE}
#using DataExplorer

plot_intro(Births, 
           ggtheme = theme_minimal(),
           title = "Overview of compiled dataset by data type")

```


This shows something of a mixed picture - with not a single column or row that is complete, looks mostly due to that Notes column, but maybe other data is missing. Looking at the missing data by variable shows some patterns worth exploring further:


```{r, echo=FALSE, layout="l-body"}
plot_missing(Births, ggtheme = theme_minimal()) + ggtitle("Overview of missing data")
```

The Notes variable which looks like a data dictionary column added to the bottom of each .txt file that was loaded:

```{r, echo=FALSE, layout="l-body"}
Births %>% 
  group_by(Notes) %>% 
  tally %>% 
  arrange(desc(n)) %>% 
  rmarkdown::paged_table(options = list(rows.print = 10))
```

Looking at Notes closer: it contains ```NA``` values, Total values, and descriptive values. The Total row look like sum values added to the end of each state, per file:

```{r}
Births %>% filter(Notes == "Total") %>% 
  select(State, Year, Notes, Births, everything()) %>% 
  rmarkdown::paged_table(options = list(rows.print = 10))

```

These totals  won't be helpful because they're based on the file year ranges, and the other notes look like data dictionary types of data, so both of these can be removed altogether by keeping only the variables that are NOT ```NA```, then drop the column altogether to remove the descriptive text.

```{r}
Births <- Births %>% filter(is.na(Notes)) %>% select(-Notes)
```
With notes removed, can get a more accurate view of missing data. The entire first file (1995 - 2002) doesn't have the same data as the other files, something to keep in mind for any future analysis.


```{r, layout="l-body-outset"}
Births %>% 
  group_by(Year) %>% 
  naniar::gg_miss_var(show_pct = TRUE, facet = Year) +
  ggtitle("% missing values, by Year")
```

```naniar::gg_miss_upset()``` is a good way to see what the relationship between missing values is:

```{r, layout="l-body-outset"}
Births %>% 
  naniar::gg_miss_upset()

```

It appears that the only missing values are from those years without the variables listed in the figure above, can do a quick check of that to verify:

```{r}
Births %>% 
  filter(Year < 2003) %>% 
  nrow()
```

Looks good (for the purposes of this exercise) - time to get on with some more fun things than data cleaning!  

## 3. Time Series Plot  

The time series plot shows a peak in births around 2007, with a trend towards the birth figures of the late 1990's returning in the mid 2010's. 

```{r, layout="l-body-outset", echo=FALSE}
Births %>% 
  group_by(Year) %>% 
  summarise(total = sum(Births)) %>% 
  ggplot(aes(x = Year, y = total)) +
  geom_line(size = 1, color = "lightgray") + 
  coord_cartesian(xlim = c(1995, 2017)) +
  ylab("Total Births")+
  scale_y_continuous(label=scales::comma,
                     limits = c(0, 4500000),
                     breaks = seq(0, 4500000, 500000)) +
  scale_x_continuous(breaks = seq(1995, 2017, 3))+
  theme_minimal(12) + 
  theme(legend.position = "none",
        axis.text.x = element_text(hjust = .75),
        plot.title = element_text(size=14, hjust = 0.5),
        plot.caption = element_text(size = 12, hjust = 1),
        axis.text.y = element_text(hjust = 1),
        plot.margin = unit(c(1,1,0.5,1), "cm")) +
  ggtitle("Total Annual US Births, 1995 - 2017")

```

Looking at each state, we see that the golden state has the highest count of births (but tied for last in basketball championship titles in 2019).

```{r, echo=FALSE, layout = "l-body-outset"}
Births %>% 
  group_by(Year, State) %>% 
  #summarise(total = sum(Births)) %>% 
  ggplot(aes(x = Year, y = Births, group = State)) +
  geom_line(size = .5, colour = "darkred") +
  gghighlight::gghighlight(max(Births) > 550000, max_highlight = 1)+
  coord_cartesian(xlim = c(1995, 2017)) +
  ylab("Total Births")+
  scale_y_continuous(label=scales::comma,
                     limits = c(0, 650000),
                     breaks = seq(0, 600000, 100000)) +
  scale_x_continuous(breaks = seq(1995, 2017, 3))+
  theme_minimal(12) + 
  theme(legend.position = "none",
        axis.text.x = element_text(hjust = .75),
        plot.title = element_text(size=14, hjust = 0.5),
        plot.caption = element_text(size = 12, hjust = 1),
        axis.text.y = element_text(hjust = 1),
        plot.margin = unit(c(1,1,0.5,1), "cm")) +
  ggtitle("Total Annual US Births by State, 1995 - 2017")
```

## 4. Linear Regression: BR impact by year & state?

Using linear regression, we can estimate how state and year influence birth rates over time:

Looking first at ```Year``` alone:

```{r}

#first filter the data before 2002 that doesn't have BR values
Births_2002_over <- Births %>% filter(Year >= 2003)

#center the Year variable
Births_2002_over <- Births_2002_over %>% mutate(Year_New = Year-2003)

#run initial model
model1 <- lm(data = Births_2002_over, formula = `Birth Rate` ~ Year_New)

tmp <- capture.output(summary(model1))

cat(tmp[c(3:4, 9:13, 16:17)], sep='\n')
```

Which can be graphically represented as:

```{r, layout = "l-body-outset"}

ggplot(Births_2002_over, aes(x = Year, y = `Birth Rate`)) +
  geom_point() +
  stat_smooth(method = "lm", formula = y~x, se=FALSE, linetype = 1) +
  scale_x_continuous(breaks = seq(1995, 2017, 3))+
  theme_minimal(12) + 
  theme(legend.position = "none",
        axis.text.x = element_text(hjust = .75),
        plot.title = element_text(size=14, hjust = 0.5),
        plot.caption = element_text(size = 12, hjust = 1),
        axis.text.y = element_text(hjust = 1),
        plot.margin = unit(c(1,1,0.5,1), "cm")) +
  ggtitle("")

```

## 4.1  Interpretation  
This linear model tells me that in 2003 there was a birth rate of 14.2, with an estimated decrease of .16 every year. The year variable is [centered](http://www.med.mcgill.ca/epidemiology/joseph/courses/EPIB-621/centered_var.pdf) around the actual years (rather than year 0), which makes the interpretation easier to interpret by essentially shifting the y-axis to the years of interest (post 2003). 

### Year and State

When accounting for State, the annual change over time sees a very marginally increased rate of decrease (from -.1596 to -.159639), and varying results for different states, with some seeing a decrease and others an increase in the birth rate over time. 

```{r}
model2 <- lm(data = Births_2002_over, formula = `Birth Rate` ~ Year_New + State)

tmp2 <- capture.output(summary(model2))

cat(tmp2[c(3:4, 9:68)], sep='\n')
```

# 5. Diffs-in-Diffs Analysis

A differences in differences (DD) analysis can be described as an analysis where outcomes are observed for two groups during two time periods,an analysis which is widespread in use although has been criticized for showing inconsistencies in resultant standard errors^[https://economics.mit.edu/files/750]. This criticism should be considered when interpreting the results of a DD analysis, but nonetheless does not mean the analysis should not be completed. 

## 5.1 Difference in Differences Case: Effect of Abstinance Policy in Three U.S. States  
I will use a a diffs-in-diffs analysis to create a plot on an abstinence only policy that was implemented in 2008 in three states: New Jersey, Georgia and Texas. The plot examines whether the parallel trends assumptions are met in the treated states versus the control states.   

<aside>
This will be completed using only states that have population denominators.
</aside>  


```{r}
Births_pop <- Births %>% 
  filter(`Total Population` > 0) 

Births_treated <- Births_pop %>% 
  mutate(
    states_treated = case_when(
      State %in% c("New Jersey", "Georgia", "Texas") ~ 1,
      TRUE ~ 0
      ),
    post2008 = Year >= 2008 
      )

```

First, the data should include only that with population data, and include an identifer differentiating the states with and without the policy, stored as Births_treated.  

When examining to see whether or not the parallel trends assumption is broken, a visual inspection works:

> "Although there is no statistical test for this assumption, visual inspection is useful when you have observations over many time points."  ^[https://www.mailman.columbia.edu/research/population-health-methods/difference-difference-estimation]

```{r, layout = "l-body-outset"}
ggplot(Births_treated, aes(Year, `Fertility Rate`, color = as.logical(states_treated))) +
  stat_summary(geom = "line") +
  geom_vline(xintercept = 2008) +
  theme_minimal(12) +
  ggtitle("Parallel Trends Assumption Test",
          subtitle = "Fertility rates in states with and without abstinance policies")
```

Based on visual inspection, the trends between both treated and non-treated states are similar, so the assumption is not violated.

## 5.2 Estimating the impact of the 2008 policy on fertility rates between states

DD is "usually implemented as an interaction term between time and treatment variables in a regression model" ^[https://www.mailman.columbia.edu/research/population-health-methods/difference-difference-estimation]; hence looking at the relationship between the treated states and the time of the intervention, the interaction term can be gathered as:

```{r}
model = lm(`Fertility Rate` ~ states_treated*post2008, data = Births_treated)
summary(model)
```

Producing a residual standard error of 7.344, indicating that the treatment *increased* by 7.344. 

In an alternative model, I can find a similar results increasing by 7.268, a similar result:

```{r}
model2 = lm(`Fertility Rate` ~ Year + states_treated + states_treated*post2008, data = Births_treated)

summary(model2)

```

Some reasons for this increase could be that along with a policy focused on abstinence comes a decrease in fundi. ng for other "family planning" and women's health supportThis would lead me to say that this isn't the best policy decision if the goal is to decrease the birth rate in a state.


